---
title: "WITSML Volve - Final"
output:
  html_document: 
    highlight: pygments
  html_notebook:
    code_folding: none
    highlight: haddock
    theme: united
---

```{r setup, include = FALSE, error=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      comment = "#>",
                      collapse = TRUE,
                      error = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.align = 'center'
                      )
```

```{r load-libraries}
# load libraries
library(xml2)
library(data.table)
library(dplyr)
```

## Retrieve WITSML file
* Using a relatively small folder of 27 MB.
* well `Norway-Statoil-NO 15_$47$_9-F-9`
* only show few of the files with their complete full name

```{r list-all-xml-files}
all_files_xml <- list.files("./witsml", recursive = TRUE, full.names = TRUE, 
                        include.dirs = TRUE, pattern = "*.xml")

# indices in R start at 1, not zero as in Python
all_files_xml[1:5]
```


```{r get-trajectory-files}
# get the file for trajectory
traj_files <- grep(pattern = "trajectory", ignore.case = TRUE, 
                   value = TRUE, x = all_files_xml)
traj_files
```

## Read the WITSML file

```{r read-xml-file}
dat <- read_xml(traj_files[1])
```

```{r xml-introspection}
# some introspection
xml_name(dat)
xml_children(dat)

# name of the child node
xml_name(xml_children(dat))
xml_name(xml_child(dat))
```

### Nodes, children and siblings

```{r first-25-nodes}
# strip default namespaces from the document
xml_ns_strip( dat )
all_nodes <- dat %>% 
    xml_find_all( '//*') %>% 
    xml_path()
    
all_nodes[1:25]                   # show only the first elements
```

```{r toal-number-of-nodes}
# get the number of elements
dat <- xml_ns_strip( dat )
noe <- dat %>% 
    xml_find_all( '//*') %>% 
    xml_path() %>% 
    length()
noe
```

```{r last-25-nodes}
# let's see the last 25
tail(all_nodes, 25)
```

## `trajectorys` root node

```{r}
# attributes for the root node
trajectorys <- xml_find_first( dat, "//trajectorys")
xml_attrs(trajectorys)
xml_name(trajectorys)
xml_siblings(trajectorys)    # no siblings because it is the root
```

## `trajectory` node

### `trajectory` summary

```{r}
# attributes of the trajectory node
trajectory <- xml_find_first( dat, "//trajectory")
xml_attrs(trajectory)
xml_name(xml_parent(trajectory))
xml_name(trajectory)
xml_name(xml_children(trajectory))
```

### `trajectory` children

```{r}
# trajectoryStation children nodes
xml_name(xml_children(xml_find_all(dat, "//trajectorys/trajectory" )))
```

At this point we don't know which node has children or dependents.

```{r}
# number of trajectoryStation children nodes
length(xml_name(xml_children(xml_find_all(dat, "//trajectorys/trajectory" ))))
```

```{r}
# another way of obtaining the names of the nodes for "trajectory"
xml_name(xml_children(xml_find_first(dat, "//trajectory")))
```

## `trajectory` childless nodes (with spec)
If we know by specification the a node does not have children then we could manually indicate the indices of the names vector.

```{r}
# name of the orphan nodes
orphan_vars <- c(1:14)   # indices

# the first 14 children of trajectory
vars14_names <- xml_name(xml_children(xml_find_first(dat, "//trajectory")))[orphan_vars]
vars14_names
```

```{r}
# get all the nodes under the root
xml_children(dat) %>%     # trajectory
    xml_children() %>%    # variables and values of the children
    xml_name()            # names of the variables only
```


```{r}
# TODO: this function just moved from below. justify it here and document
# get the names of variables under a node
get_variables_under_node <- function(xml_dat, node) {
    xpath <- paste("//", node)
    xml_find_all(xml_dat, xpath) %>% 
    xml_children() %>% 
    xml_name() %>% 
    unique()
}
```


### functions `how_many_children`, `have_children`

We could check if the nodes under `trajectory` have children.

```{r function-how-many-children}
how_many_children <- function(xml_dat, node) {
  vars_vector <- vector("integer")
  var_names <- get_variables_under_node(xml_dat, node)
  i <- 1
  for (var in var_names) {
    xpath <- paste("//", node, "/", var)  
    num_children <- max(xml_length(xml_find_all(xml_dat, xpath)))
    vars_vector[i] <- num_children
    names(vars_vector)[i] <- var
    # cat(i, var, vars_vector[i], "\n")
    i <- i + 1
  } 
  vars_vector
}

how_many_children(dat, "trajectory")
```

```{r function-have-children}
have_children <- function(xml_dat, node) {
  how_many <- how_many_children(xml_dat, node)
  how_many[how_many > 0]
}

have_no_children <- function(xml_dat, node) {
  how_many <- how_many_children(xml_dat, node)
  how_many[how_many == 0]
}
```

### childless nodes
Now we can easily find what nodes are childless

```{r find-children-trajectory-childless}
have_no_children(dat, "trajectory")
names(have_no_children(dat, "trajectory"))
```


## `//trajectory/trajectoryStation`

```{r}
# names of the children for "trajectoryStation"
xml_name(xml_children(xml_find_first( dat, "//trajectory/trajectoryStation")))
```

This yields the same result:

```{r}
# names of the nodes for "trajectoryStation"
xml_name(xml_children(xml_find_first(dat, "//trajectoryStation")))
```

```{r}
# find attributes of the first element of trajectoryStation found
trajectoryStation <- xml_find_first(dat, "//trajectoryStation")
xml_attrs(x = trajectoryStation)
```

### number of measurement stations

```{r}
# find all observations for dTimStn
trajectoryStation.dTimStn <- xml_find_all(dat, "//trajectoryStation/dTimStn")
trajectoryStation.dTimStn

# we end up finding a way to calculate the number of trajectory stations
length(trajectoryStation.dTimStn)
```

```{r}
# name of the dependent nodes of "trajectoryStation"
xml_name(xml_children(xml_find_first(dat, "//trajectoryStation")))
```

```{r}
# get the attributes for trajectoryStation
xml_attrs(x = trajectoryStation)

# we get only the "uid" attribute
```

```{r}
# get the value of the attribute we found
xml_attr(x = trajectoryStation, attr = "uid")
```

## commonData: names and values

There are two nodes with the same name:
* `//trajectory/commonData`
* `//trajectoryStation/commonData`

### //trajectory/commonData
```{r}
# get the subnodes for //trajectory/commonData
unique(xml_name(xml_children(xml_find_all(dat, "//trajectory/commonData"))))
```

```{r}
# number of subnodes
max(xml_length(xml_find_all(dat, "//trajectory/commonData")))
```

```{r}
# values for the first macthing node
xml_text((xml_find_first(dat, "//trajectory/commonData" ) ) )
```

### //trajectoryStation/commonData
```{r}
# get the subnodes for //trajectoryStation/commonData
unique(xml_name(xml_children(xml_find_all(dat, "//trajectoryStation/commonData" ) ) ))
```

```{r}
# number of subnodes for //trajectoryStation/commonData
# we use max() for the case it returns a vector with multiple nodes and lengths
max(xml_length(xml_find_all(dat, "//trajectoryStation/commonData")))
```

```{r}
# values for the first matching node
xml_text((xml_find_first(dat, "//trajectoryStation/commonData" ) ) )
```

# Create the dataframes

## trajectory

```{r}
# get all attributes for trajectory node
# we try with datatable and dataframe
trajectory <- xml_find_first( dat, "//trajectory")
trajectory <- xml_attrs(trajectory)

trajectory_dt <- data.table(t(trajectory))
trajectory_dt

trajectory_df <- data.frame(t(trajectory), stringsAsFactors = FALSE)
trajectory_df
```

```{r}
# get values for all the attributes of the trajectory node
# using magrittr
xml_find_first( dat, "//trajectory") %>% 
    xml_attrs() %>% 
    t() %>% 
    data.frame(stringsAsFactors = FALSE)
```

## Siblings of `trajectoryStation`

```{r}
# get all the trajectory variables, with and without descendants
trajectory_children <- xml_name(xml_children(xml_find_first(dat, "//trajectory")))

# get only those variables without descendants. hnc: have no children
traj_hnc_names <- names(have_no_children(dat, "trajectory"))
traj_hnc_idx <- which(trajectory_children %in% traj_hnc_names)

traj_hnc <- xml_text(xml_children(xml_find_first( dat, "//trajectory")))[traj_hnc_idx]
names(traj_hnc) <- traj_hnc_names
traj_hnc

# dataframe and datatable
siblings_df <- data.frame(t(traj_hnc), stringsAsFactors = FALSE)
siblings_df

siblings_dt <- data.table(t(traj_hnc))
siblings_dt
```

### combine trajectory and siblings in a one-row dataframe

```{r}
cbind(trajectory_dt, siblings_dt)
cbind(trajectory_df, siblings_df)
```


## `trajectoryStation` measurements

### dataframe for trajectoryStation, uid

```{r rows.print=25}
# get values for uid attribute of trajectoryStation
# these are the well ids
tS.uid <- dat %>% 
    xml_find_all("//trajectoryStation") %>% 
    xml_attr("uid")

tS.uid_dt <- data.table(uid = tS.uid)
tS.uid_df <- data.frame(uid = tS.uid, stringsAsFactors = FALSE)

tS.uid_dt
tS.uid_df
```

We could use this method to create a function that finds the number of observations for the `trajectoryStation`:

```{r function-number-tS}
# get the number of rows for a parent node
#' Get the number of rows for a parent node.
#'
#' @param xml_dat the xml document
#' @param parent_node a node without the two forward slashes
#' @param attribute the attribute of the node if available
#'
#' @return
#' @export
#'
#' @examples
get_numrows_parent_node <- function(xml_dat, parent_node, attribute) {
    # TODO: validate if the node has an attribute
    xml_dat %>% 
    xml_find_all(paste("//", parent_node)) %>%
    xml_attr("uid") %>%
    length()
}

# exercise the function
parent <- "trajectoryStation"
attrib <- "uid"
get_numrows_parent_node(dat, parent, attrib)

```


## measurement stations or `trajectoryStation`

```{r}
# using xml_children
# we also get commonData which has children
trajectoryStation_all_names <- xml_name(xml_children(xml_find_all(dat, 
                                                            "//trajectoryStation")))
trajectoryStation_all_names <- unique(trajectoryStation_all_names)
trajectoryStation_all_names
```


```{r}
# get the number of columns by name
# commonData is excluded (but we know that in advance)
no_commonData <- which(trajectoryStation_all_names %in% c("commonData"))

# get rid of commonData since it has children
trajectoryStation_names <- trajectoryStation_all_names[-no_commonData] # exclude
trajectoryStation_names
```

### draft of a future function

```{r draft-function}
# first non-automated way of getting values for the all the trajectoryStation nodes
# there are 19 variables under trajectoryStation, not including commonData which
# was manually removed
xml_dat <- dat
node <- "trajectoryStation"
max_obs <- get_numrows_parent_node(dat, "trajectoryStation", attribute = "uid")
var_names <- trajectoryStation_names  # names of the variables in a vector
li_vars <- vector("list")             # vector of list
for (var in var_names) {              # iterate through all the variables
    xpath <- paste("//", node, "/", var)  # form the xpath
    value_xpath <- xml_text(xml_find_all(xml_dat, xpath)) # get all the values
    vx <- value_xpath                                  # make it a shorter name
    # if the variables are all not present, add NA. max=25
    if (length(vx) < max_obs) vx <- c(rep(NA, max_obs - length(vx)), vx)
    li_vars[[var]] <- vx
}
tS_df <- as.data.frame(li_vars, stringsAsFactors = FALSE)
tS_df
```

```{r function-get-variables}
# get the names of variables under a node
get_variables_under_node <- function(xml_dat, node) {
    xpath <- paste("//", node)
    xml_find_all(xml_dat, xpath) %>% 
    xml_children() %>% 
    xml_name() %>% 
    unique()
}

tS.cD_names <- get_variables_under_node(dat, "trajectoryStation/commonData")
tS.cD_names
```


```{r}
# get variables under trajectoryStation
# detect what variables are standalone and which ones have children
how_many_children(dat, "trajectoryStation")
```

There are two variables under `trajectoryStation` that have descendants.

```{r function-nodes_as_df}
#' Converts children of a node and their values to a dataframe.
#' Receives a node (do not add '//'), creates a vector with the variables under
#' the node, iterates through each of the variables, fills uneven rows with NAs.
#' It will skip a child node that contains children.
#'
#' @param xml_dat a xml document
#' @param node a node of the form "trajectoryStation/dTimStn". No need to add "//"
#' @param max_obs 
nodes_as_df <- function(xml_dat, node, max_obs) {
    li_vars <- vector("list")             # vector of list
    var_names <- get_variables_under_node(xml_dat, node)
    for (var in var_names) {              # iterate through all the variables
        xpath <- paste("//", node, "/", var)  # form the xpath
        num_children <- max(xml_length(xml_find_all(xml_dat, xpath)))
        if (num_children == 0) {  # skip if the node has children
            value_xpath <- xml_text(xml_find_all(xml_dat, xpath)) # get all the values
            vx <- value_xpath                                  # make it a shorter name
            # if the variables are all not present, add NA. max=25
            if (length(vx) < max_obs) vx <- c(rep(NA, max_obs - length(vx)), vx)
            li_vars[[var]] <- vx
        }
    }
    as.data.frame(li_vars, stringsAsFactors = FALSE)
}

# using function get_numrows_parent_node()
num_trajectoryStation <- get_numrows_parent_node(dat, "trajectoryStation", 
                                                 attribute = "uid")
tS.trajectoryStation_df <- nodes_as_df(dat, "trajectoryStation", num_trajectoryStation)
tS.trajectoryStation_df
```


### //trajectoryStation/commonData

```{r}
# find children for commonData
xpath <- "//trajectoryStation/commonData"
trajectoryStation.cD_names <- xml_name(xml_children(xml_find_all(dat, xpath)))
trajectoryStation.cD_names <- unique(trajectoryStation.cD_names)
trajectoryStation.cD_names
```

```{r}
# cascading way
xpath <- "//trajectoryStation/commonData"
xml_find_all(dat, xpath) %>% 
    xml_children() %>% 
    xml_name() %>% 
    unique()
```

### dataframe for trajectoryStation/commonData

```{r}
# get the nodes under trajectoryStation/commonData
tS.cD_df <- nodes_as_df(xml_dat, node = "trajectoryStation/commonData", max_obs = 25)
tS.cD_df
```

### dataframe of combined trajectoryStation

```{r rows.print=25}
# combine all dataframes to make up trajectoryStation dataframe
  # tS.uid_dt: trajectoryStation attributes
  # tS_df: trajectoryStation data
  # tS.cD_df: commonData
trajectoryStation_df <- cbind(tS.uid_dt, tS_df, tS.cD_df)
trajectoryStation_df
```

```{r}
dim(trajectoryStation_df)
names(trajectoryStation_df)

# TODO: decompose corUsed and add it to the dataframe
```


# Notes

* For the first row of `trajectoryStation`, which corresponds to the `TIE_POINT`, make all the empty variables (or not present) as NA. The `data.table` function works very well even out those variables that are incomplete but fill them with default values. We see that the  `trajectoryStation` variables are only 10 without counting `commonData`.

* In upcoming versions take care of coercing the variables to their corresponding types. By default, in this example, we've got all the variables as character. Because we know in advance the data types, we could use **R** package `readr` to do that.

* As we get familiar with the WITSML hierarchies, we could start using loops or `apply` functions to convert a tree to a dataframe.

* Functions can be implemented later to get the number of trajectory stations, find which `trajectoryStation` does not have its complete set of variables, or extract a particular trajectory measurement.

* In this example, we used only one well. That's why we obtained only one trajectory file. Other wells could have more than one trajectory file. Later we could implement a function that scans all the folders and generates an statistic of the number of folders and number of files per well.* For the first row of `trajectoryStation`, which corresponds to the `TIE_POINT`, make all the empty variables (or not present) as NA. The `data.table` function works very well even out those variables that are incomplete but fill them with default values. We see that the  `trajectoryStation` variables are only 10 without counting `commonData`.

* In upcoming versions take care of coercing the variables to their corresponding types. By default, in this example, we've got all the variables as character. Because we know in advance the data types, we could use **R** package `readr` to do that.

* As we get familiar with the WITSML hierarchies, we could start using loops or `apply` functions to convert a tree to a dataframe.

* Functions can be implemented later to get the number of trajectory stations, find which `trajectoryStation` does not have its complete set of variables, or extract a particular trajectory measurement.

* In this example, we used only one well. That's why we obtained only one trajectory file. Other wells could have more than one trajectory file. Later we could implement a function that scans all the folders and generates an statistic of the number of folders and number of files per well.

