---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r}
# load libraries
library(xml2)
library(data.table)
library(dplyr)
```

## Retrieve WITSML file
* Using a relatively small folder of 27 MB.
* well `Norway-Statoil-NO 15_$47$_9-F-9`
* only show few of the files with their complete full name

```{r}
all_files_xml <- list.files("./witsml", recursive = TRUE, full.names = TRUE, 
                        include.dirs = TRUE, pattern = "*.xml")

# indices in R start at 1, not zero as in Python
all_files_xml[1:5]
```


```{r}
# get the file for trajectory
traj_files <- grep(pattern = "trajectory", ignore.case = TRUE, 
                   value = TRUE, x = all_files_xml)
traj_files
```

## Read the WITSML file

```{r}
dat <- read_xml(traj_files[1])
```

```{r}
# some introspection
xml_name(dat)
xml_children(dat)

# name of the child node
xml_name(xml_children(dat))
xml_name(xml_child(dat))
```

## Nodes, children and siblings

```{r}
# strip default namespaces from the document
xml_ns_strip( dat )
dat %>% 
    xml_find_all( '//*') %>% 
    xml_path() %>% 
    .[1:50]                   # show only the first elements
```

```{r}
# get the number of elements
xml_ns_strip( dat )
noe <- dat %>% 
    xml_find_all( '//*') %>% 
    xml_path() %>% 
    length()
noe
```

### `trajectorys` root node

```{r}
# attributes for the root node
trajectorys <- xml_find_first( dat, "//trajectorys")
xml_attrs(trajectorys)
xml_name(trajectorys)
xml_siblings(trajectorys)    # no siblings because it is the root
```

### `trajectory` child

```{r}
# attributes of the trajectory node
trajectory <- xml_find_first( dat, "//trajectory")
xml_attrs(trajectory)
xml_name(xml_parent(trajectory))
xml_name(trajectory)
xml_name(xml_children(trajectory))
```

### `trajectory` children

```{r}
# trajectoryStation children nodes
xml_name(xml_children(xml_find_all(dat, "//trajectorys/trajectory" )))
```

```{r}
# number of trajectoryStation children nodes
length(xml_name(xml_children(xml_find_all(dat, "//trajectorys/trajectory" ))))
```

```{r}
# another way of obtaining the names of the nodes for "trajectory"
xml_name(xml_children(xml_find_first(dat, "//trajectory")))
```

## `trajectory` childless nodes
If we know by specification the a node does not have children then we could manually indicate the indices of the names vector.

```{r}
# name of the orphan nodes
orphan_vars <- c(1:14)   # indices

# the first 14 children of trajectory
vars14_names <- xml_name(xml_children(xml_find_first(dat, "//trajectory")))[orphan_vars]
vars14_names
```

```{r}
# get all the nodes under the root
xml_children(dat) %>%     # trajectory
    xml_children() %>%    # variables and values of the children
    xml_name()            # names of the variables only
```


## `//trajectory/trajectoryStation`

```{r}
# names of the children for "trajectoryStation"
xml_name(xml_children(xml_find_first( dat, "//trajectory/trajectoryStation")))
```

This yields the same result:

```{r}
# names of the nodes for "trajectoryStation"
xml_name(xml_children(xml_find_first(dat, "//trajectoryStation")))
```

```{r}
# find attributes of the first element of trajectoryStation found
trajectoryStation <- xml_find_first(dat, "//trajectoryStation")
xml_attrs(x = trajectoryStation)
```

### number of measurement stations

```{r}
# find all observations for dTimStn
trajectoryStation.dTimStn <- xml_find_all(dat, "//trajectoryStation/dTimStn")
trajectoryStation.dTimStn

# we end up finding a way to calculate the number of trajectory stations
length(trajectoryStation.dTimStn)
```

```{r}
# name of the dependent nodes of "trajectoryStation"
xml_name(xml_children(xml_find_first(dat, "//trajectoryStation")))
```

```{r}
# get the attributes for trajectoryStation
xml_attrs(x = trajectoryStation)

# we get only the "uid" attribute
```

```{r}
# get the value of the attribute we found
xml_attr(x = trajectoryStation, attr = "uid")
```

## commonData: names and values

There are two nodes with the same name:
* `//trajectory/commonData`
* `//trajectoryStation/commonData`

### //trajectory/commonData
```{r}
# get the subnodes for //trajectory/commonData
unique(xml_name(xml_children(xml_find_all(dat, "//trajectory/commonData"))))
```

```{r}
# number of subnodes
max(xml_length(xml_find_all(dat, "//trajectory/commonData")))
```

```{r}
# values for the first macthing node
xml_text((xml_find_first(dat, "//trajectory/commonData" ) ) )
```

### //trajectoryStation/commonData
```{r}
# get the subnodes for //trajectoryStation/commonData
unique(xml_name(xml_children(xml_find_all(dat, "//trajectoryStation/commonData" ) ) ))
```

```{r}
# number of subnodes for //trajectoryStation/commonData
# we use max() for the case it returns a vector with multiple nodes and lengths
max(xml_length(xml_find_all(dat, "//trajectoryStation/commonData")))
```

```{r}
# values for the first matching node
xml_text((xml_find_first(dat, "//trajectoryStation/commonData" ) ) )
```

# Create the dataframes

## trajectory

```{r}
# get all attributes for trajectory node
# we try with datatable and dataframe
trajectory <- xml_find_first( dat, "//trajectory")
trajectory <- xml_attrs(trajectory)

trajectory_dt <- data.table(t(trajectory))
trajectory_dt

trajectory_df <- data.frame(t(trajectory), stringsAsFactors = FALSE)
trajectory_df
```

```{r}
# get all attributes for the trajectory node
# using magrittr
xml_find_first( dat, "//trajectory") %>% 
    xml_attrs() %>% 
    t() %>% 
    data.frame(stringsAsFactors = FALSE)
```

## Siblings of `trajectoryStation`

> Maybe we should try with a different name for `orphan_vars`.

```{r}
# this is the manual way of obtaining the table
# values of the orphan nodes
orphan_vars <- c(1:14)   # indices if we know the specs in advance

vars14_values <- xml_text(xml_children(xml_find_first( dat, "//trajectory")))[orphan_vars]
names(vars14_values) <- vars14_names

# dataframe and datatable
orphans_df <- data.frame(t(vars14_values), stringsAsFactors = FALSE)
orphans_df

orphans_dt <- data.table(t(vars14_values))
orphans_dt
```

```{r eval=FALSE}
# get the names of the siblings of trajectoryStation
# tS: trajectoryStation
orphan_vars_ix <- c(1:14)                          # this is not the best solution
names_siblings_tS <- xml_find_first(dat, "//trajectory") %>% 
    xml_children() %>% 
    xml_name() %>% 
    .[orphan_vars_ix]

names_siblings_tS
```

```{r}
# using xml_siblings
trajectoryStation_names <- xml_name(xml_siblings(xml_find_all(dat, "//trajectoryStation")))
trajectoryStation_names <- unique(trajectoryStation_names)
trajectoryStation_names
```

```{r}
# we don't want trajectoryStation nor commonData
not_these_cols <- which(trajectoryStation_names %in% c("trajectoryStation", "commonData"))
orphan_vars <- trajectoryStation_names[-c(not_these_cols)]
orphan_vars
```


```{r}
# get the values for the siblings of trajectory
values_siblings_tS <- xml_find_first(dat, "//trajectory") %>% 
    xml_children() %>% 
    xml_text() %>% 
    .[which(trajectoryStation_names %in% orphan_vars)]

siblings_tS <- values_siblings_tS
siblings_tS
```


```{r}
# assign names to vectors with values
names(siblings_tS) <- names_siblings_tS
siblings_tS
```

```{r}
# convert the vectors to table
siblings_dt <- siblings_tS %>% 
    t() %>% 
    data.table()

siblings_df <- siblings_tS %>% 
    t() %>% 
    data.frame(stringsAsFactors = FALSE)

siblings_dt
siblings_df
# siblings_dt replaces orphans_dt
```


## combine trajectory and siblings in a one-row dataframe

```{r}
cbind(trajectory_dt, siblings_dt)
cbind(trajectory_df, siblings_df)
```


## trajectoryStation

### dataframe for trajectoryStation, uid

```{r rows.print=25}
# get values for uid attribute of trajectoryStation
# these are the well ids
tS.uid <- dat %>% 
    xml_find_all("//trajectoryStation") %>% 
    xml_attr("uid")

tS.uid_dt <- data.table(uid = tS.uid)
tS.uid_dt

tS.uid_df <- data.frame(uid = tS.uid, stringsAsFactors = FALSE)
tS.uid_df
```

```{r rows.print=25}
# get values for uid attribute of trajectoryStation
# these are the well ids
tS.uid <- dat %>% 
    xml_find_all("//trajectoryStation") %>% 
    xml_attr("uid")
tS.uid_df <- data.frame(uid = tS.uid, stringsAsFactors = FALSE)
tS.uid_df
```

We could use this method to create a function that finds the number of observations for the `trajectoryStation`:

```{r function-number-tS}
# get the number of rows for a parent node
get_numrows_parent_node <- function(xml_dat, parent_node, attribute) {
    xml_dat %>% 
    xml_find_all(paste("//", parent_node)) %>%
    xml_attr("uid") %>%
    length()
}

# exercise the function
parent <- "trajectoryStation"
attrib <- "uid"
get_numrows_parent_node(dat, parent, attrib)

```


## measurement stations or trajectoryStation

```{r}
# using xml_children
# we also get commonData
trajectoryStation_all_names <- xml_name(xml_children(xml_find_all(dat, 
                                                            "//trajectoryStation")))
trajectoryStation_all_names <- unique(trajectoryStation_all_names)
trajectoryStation_all_names
```


```{r}
# get the number of columns by name
# commonData is excluded
no_commonData <- which(trajectoryStation_all_names %in% c("commonData"))

# get rid of commonData since it has children
trajectoryStation_names <- trajectoryStation_all_names[-no_commonData] # exclude
trajectoryStation_names
```


```{r}
# there are 19 variables under trajectoryStation, not including commonData
xml_dat <- dat
node <- "trajectoryStation"
max_obs <- get_numrows_parent_node(dat, "trajectoryStation", attribute = "uid")
var_names <- trajectoryStation_names  # names of the variables in a vector
li_vars <- vector("list")             # vector of list
for (var in var_names) {              # iterate through all the variables
    xpath <- paste("//", node, "/", var)  # form the xpath
    value_xpath <- xml_text(xml_find_all(xml_dat, xpath)) # get all the values
    vx <- value_xpath                                  # make it a shorter name
    # if the variables are all not present, add NA. max=25
    if (length(vx) < max_obs) vx <- c(rep(NA, max_obs - length(vx)), vx)
    li_vars[[var]] <- vx
}
tS_df <- as.data.frame(li_vars, stringsAsFactors = FALSE)
tS_df
```

```{r function-get-variables}
# get the names of variables under a node
get_variables_under_node <- function(xml_dat, node) {
    xpath <- paste("//", node)
    xml_find_all(xml_dat, xpath) %>% 
    xml_children() %>% 
    xml_name() %>% 
    unique()
}

tS.cD_names <- get_variables_under_node(dat, "trajectoryStation/commonData")
tS.cD_names
```

```{r}
# detect what variables are standalone and which ones have children
# get variables under trajectoryStation
var_names <- get_variables_under_node(dat, node = "trajectoryStation")
node = "trajectoryStation"

for (var in var_names) {
    xpath <- paste("//", node, "/", var)  
    num_children <- max(xml_length(xml_find_all(dat, xpath)))
    cat(var, num_children, "\n")
}    
```



```{r function-nodes_as_df}
#' Converts children of a node and their values to a dataframe.
#' Receives a node (do not add '//'), creates a vector with the variables under
#' the node, iterates through each of the variables, fills uneven rows with NAs.
#' It will skip a child node that contains children.
#'
#' @param xml_dat a xml document
#' @param node a node of the form "trajectoryStation/dTimStn". No need to add "//"
#' @param max_obs 
nodes_as_df <- function(xml_dat, node, max_obs) {
    li_vars <- vector("list")             # vector of list
    var_names <- get_variables_under_node(xml_dat, node)
    for (var in var_names) {              # iterate through all the variables
        xpath <- paste("//", node, "/", var)  # form the xpath
        num_children <- max(xml_length(xml_find_all(xml_dat, xpath)))
        if (num_children == 0) {  # skip if the node has children
            value_xpath <- xml_text(xml_find_all(xml_dat, xpath)) # get all the values
            vx <- value_xpath                                  # make it a shorter name
            # if the variables are all not present, add NA. max=25
            if (length(vx) < max_obs) vx <- c(rep(NA, max_obs - length(vx)), vx)
            li_vars[[var]] <- vx
        }
    }
    as.data.frame(li_vars, stringsAsFactors = FALSE)
}

# using function get_numrows_parent_node()
num_trajectoryStation <- get_numrows_parent_node(dat, "trajectoryStation", 
                                                 attribute = "uid")
tS.trajectoryStation_df <- nodes_as_df(dat, "trajectoryStation", num_trajectoryStation)
tS.trajectoryStation_df
```

```{r}

```


```{r}
# exercise the function
var_names <- trajectoryStation_names    # names of the variables in a vector
# max_obs <- get_numrows_parent_node(dat, parent, attrib)
rows_trajectoryStation <- 25  # hard value
nodes_as_df(dat, "trajectoryStation", rows_trajectoryStation)
```

```{r eval=FALSE}
# the function above replaces all these assignments
# not the best solution
node <- "trajectoryStation"
dTimStn <- xml_text( xml_find_all( dat, paste("//", node, "/", "dTimStn")))
typeTrajStation <- xml_text( xml_find_all( dat, paste("//", node, "/", "typeTrajStation")))
md <- xml_text(xml_find_all(dat, paste("//", node, "/", "md")))
tvd <- xml_text(xml_find_all(dat, paste("//", node, "/", "tvd")))
incl <- xml_text(xml_find_all(dat, paste("//", node, "/", "incl")))
azi <- xml_text(xml_find_all(dat, paste("//", node, "/", "azi")))
dispNs <- xml_text(xml_find_all(dat, paste("//", node, "/", "dispNs")))
dispEw <- xml_text(xml_find_all(dat, paste("//", node, "/", "dispEw")))
vertSect <- xml_text(xml_find_all(dat, paste("//", node, "/", "vertSect")))
dls <- xml_text(xml_find_all(dat, paste("//", node, "/", "dls")))
rateTurn <- xml_text(xml_find_all(dat, paste("//", node, "/", "rateTurn")))
rateBuild <- xml_text(xml_find_all(dat, paste("//", node, "/", "rateBuild")))
gravAccelCorUsed <- xml_text(xml_find_all(dat, paste("//", node, "/", "gravAccelCorUsed")))
magXAxialCorUsed <- xml_text(xml_find_all(dat, paste("//", node, "/", "magXAxialCorUsed")))
sagCorUsed <- xml_text(xml_find_all(dat, paste("//", node, "/", "sagCorUsed")))
magDrlstrCorUsed <- xml_text(xml_find_all(dat, paste("//", node, "/", "magDrlstrCorUsed")))
statusTrajStation <- xml_text(xml_find_all(dat, paste("//", node, "/", "statusTrajStation")))
corUsed <- xml_text(xml_find_all(dat, paste("//", node, "/", "corUsed")))

tS.trajectoryStation_dt <- data.table(dTimStn, typeTrajStation, md, tvd, incl, azi,
                                 dispNs, dispEw, vertSect, dls, 
                                 rateTurn, rateBuild, gravAccelCorUsed, magXAxialCorUsed,
                                 sagCorUsed, magDrlstrCorUsed, statusTrajStation,
                                 corUsed
                                 )
tS.trajectoryStation_dt
```

```{r}
# find children for commonData
xpath <- "//trajectoryStation/commonData"
trajectoryStation.cD_names <- xml_name(xml_children(xml_find_all(dat, xpath)))
trajectoryStation.cD_names <- unique(trajectoryStation.cD_names)
trajectoryStation.cD_names
```

```{r}
# cascading way
xpath <- "//trajectoryStation/commonData"
xml_find_all(dat, xpath) %>% 
    xml_children() %>% 
    xml_name() %>% 
    unique()
```



```{r}
# get the nodes under trajectoryStation/commonData
tS.cD_df <- nodes_as_df(xml_dat, node = "trajectoryStation/commonData", max_obs = 25)
tS.cD_df
```

```{r}
# combine all dataframes to make up trajectoryStation dataframe
trajectoryStation_df <- cbind(tS.uid_dt, tS_df, tS.cD_df) %>% 
    select(-commonData)
trajectoryStation_df
```

```{r}
# get the values for the trajectoryStation/commonData variables
# using function get_numrows_parent_node()
num_trajectoryStation <- get_numrows_parent_node(dat, "trajectoryStation", 
                                                 attribute = "uid")

tS.commonData_df <- nodes_as_df(dat, "trajectoryStation/commonData", num_trajectoryStation)
tS.commonData_df
```



```{r eval=FALSE}
# the function above replaces this manual assignment
# cD: commonData node under trajectoryStation
node <- "trajectoryStation/commonData"
cD.sourceName <- xml_text(xml_find_all(dat, paste("//", node, "/", "sourceName")))
cD.dTimCreation <- xml_text(xml_find_all(dat, paste("//", node, "/", "dTimCreation")))
cD.dTimLastChange <- xml_text(xml_find_all(dat, paste("//", node, "/", "dTimLastChange")))
cD.itemState <- xml_text(xml_find_all(dat, paste("//", node, "/", "itemState")))
cD.priv_customData <- xml_text(xml_find_all(dat, paste("//", node, "/", "priv_customData")))

tS.commonData_dt <- data.table(sourceName = cD.sourceName,
                            dTimCreation = cD.dTimCreation,
                            dTimLastChange = cD.dTimLastChange,
                            itemState = cD.itemState,
                            priv_customData = cD.priv_customData
                            )
tS.commonData_dt
```



```{r eval=FALSE}
# binding the two dataframes: trajectoryStation, commonData
cbind(tS.uid_df, tS.trajectoryStation_df, tS.commonData_df)
```



# Notes

* For the first row of `trajectoryStation`, which corresponds to the `TIE_POINT`, make all the empty variables (or not present) as NA. The `data.table` function works very well even out those variables that are incomplete but fill them with default values. We see that the  `trajectoryStation` variables are only 10 without counting `commonData`.

* In upcoming versions take care of coercing the variables to their corresponding types. By default, in this example, we've got all the variables as character. Because we know in advance the data types, we could use **R** package `readr` to do that.

* As we get familiar with the WITSML hierarchies, we could start using loops or `apply` functions to convert a tree to a dataframe.

* Functions can be implemented later to get the number of trajectory stations, find which `trajectoryStation` does not have its complete set of variables, or extract a particular trajectory measurement.

* In this example, we used only one well. That's why we obtained only one trajectory file. Other wells could have more than one trajectory file. Later we could implement a function that scans all the folders and generates an statistic of the number of folders and number of files per well.* For the first row of `trajectoryStation`, which corresponds to the `TIE_POINT`, make all the empty variables (or not present) as NA. The `data.table` function works very well even out those variables that are incomplete but fill them with default values. We see that the  `trajectoryStation` variables are only 10 without counting `commonData`.

* In upcoming versions take care of coercing the variables to their corresponding types. By default, in this example, we've got all the variables as character. Because we know in advance the data types, we could use **R** package `readr` to do that.

* As we get familiar with the WITSML hierarchies, we could start using loops or `apply` functions to convert a tree to a dataframe.

* Functions can be implemented later to get the number of trajectory stations, find which `trajectoryStation` does not have its complete set of variables, or extract a particular trajectory measurement.

* In this example, we used only one well. That's why we obtained only one trajectory file. Other wells could have more than one trajectory file. Later we could implement a function that scans all the folders and generates an statistic of the number of folders and number of files per well.


